---
title: "Intra-Family Variability in Non-Cognitive Skills of Siblings During Adolescence and Youth: Insights From the Household Random Intercept Terms"
format:
  tandf-pdf:
    keep-tex: true  
  tandf-html: default
  docx: default
author:
  - name: Garen Avanesian
    affiliations:
      - ref: SFedU
    orcid: 0000-0002-0328-637X
    email: avanesyan@sfedu.ru
  - name: Ksenia Rozhkova
    affiliations:
      - ref: HSE
    orcid: 0000-0003-1497-5733
    email: krozhkova@hse.ru
affiliations:
  - id: SFedU
    name: Southern Federal University
    department: Future Skills Research Lab
    city: Rostov-on-Don
    country: Russia
  - id: HSE
    name: Higher School of Economics
    department: Labor Market Studies Lab
    city: Moscow
    country: Russia
abstract: |
  
keywords: 
  - non-cognitive skills 
  - within-family variability
  - big five personality traits
  - family-niche model
  - socio-economic status 
  - mixed-effects models
bibliography: references.bib  
biblio-style: bwl-FU
citeproc: true
csl: "apa-with-abstract.csl"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)


```

```{r}
main_folder = "~/Documents/GitHub/Thesis"

setwd(main_folder)

source(file.path(main_folder, "02_codes/libraries.R"))
source(file.path(main_folder, "02_codes/IntraFamVar/01_family_membership.R"))
source(file.path(main_folder, "02_codes/IntraFamVar/02_data_select.R"))
source(file.path(main_folder, "02_codes/IntraFamVar/03_regressions.R"))

```

## Motivation

A family is widely recognized as the primary agent of socialization, serving as the foundation where children internalize values, acquire social norms, and develop their unique personalities. It is within the family environment that the seeds of personality traits, or, as they are called in economics, non-cognitive skills, find their nature and nurture. Only after this foundational stage do institutional factors, such as education and labor markets, come into play as individuals navigate the various transitions of the life course, from childhood to adolescence, youth, and adulthood.

While substantial research in economics and other related disciplines has examined how non-cognitive skills influence key socio-economic outcomes, ranging from educational attainment and labor market performance to health and well-being, the mechanisms underlying the development of these skills remain poorly understood. Specifically, economic research did not go further than acknowledging that these traits are function of both hereditary mechanisms, as well as social milieu <!--# reference -->. With that respect, there is limited understanding of how family dynamics and individual-specific experiences contribute to the variability in non-cognitive skills among individuals raised in one household, i.e., those who share the same genetic traits and factors of socio-economic environment.

This study seeks to address this gap by examining the within-family variability in non-cognitive skills between siblings. In addition to the common genetics, by sharing the same parents the siblings do not vary in parental education, occupation, household income, culture, religion, values and norms, and other family-specific characteristics. In other words, the confounding of these characteristics is thus neglected. This allows us to isolate the unique individual behaviors that contribute to the observed differences in non-cognitive skills among siblings.

By focusing on the Big Five traits, which are widely recognized as the most robust and comprehensive framework for understanding individual differences in personality, the study aims to uncover the mechanisms behind these skills and explain the observed differences among individuals who live in the same household and are blood-related siblings. This approach provides a nuanced understanding of the interplay between family environment and individual agency in the development of non-cognitive skills, offering insights with implications for both developmental theory and policy interventions.

To better estimate the influence of observed and unobserved family factors on non-cognitive skills, it is crucial to narrow the focus from the general population to a specific age cohort. While most of the analyses of non-cognitive skills focus either on children or on youth or on working population specifically, the current research switches the angle. It is fairly out of discussion that childhood is primarily associated with developmental processes, which means that loosing on this group would result in potentially missing important changes in the ways family environment affects non-cognitive skills. However, it is also critical to estimate how non-cognitive skills affect socio-economic outcomes during youth, controlling for the effects of the family bacground and isolating the family-level cgaracteristics in the analysis of how non cognitive skills affect continuity of education and employment.

For these reasons, this study focuses on young women and men aged 10 to 29, an age group where transitional effects are most evident, and the influence of major life milestones on non-cognitive skills can be clearly observed.

This study is guided by several research questions. The first one is: **What is the role of family factors in explaining individual variation in non-cognitive skills of siblings?** In other words, the purpose of study is to estimate the effect of individually variant characteristics on variation of non-cognitive skills in siblings who share the same family characteristics. To support this primary inquiry, several supplementary analyses are conducted, addressing the following questions:

-   What are the effects of demographic factors, such as age, sex, and birth order (e.g., being a first-born child), on non-cognitive skills?

-   Do childhood relationships with parents have lasting effects on the development of non-cognitive skills?

-   To what extent does being in adverse health status during youth influence non-cognitive skills?

Second question twists the prespective. it aims to estimate what is the effect of non-cognitive skills on continuity of education (choice to obtain tertiary degree) and employment, controlling for the family background. The question wiht that respect is: Do non-cognitive skills poroduce statistically significant effects on decesion to obtain higher edu degree and finding a job for the siblings during the school-to-work transition age (15-29) controlling for the shared family bacground? In other words this angle allows to estimate if non-cognitive skills still play a role in these socio economic outcomes ansd their effects identified in literature previousely do not occur due to the advantage of certain children due to socio-economic characteristics of family, such as social capital, parental education, occupation, and household income.

## Previous Evidence

Historically, much research on the differences in personality in a within-family design attributes them to the birth order and was carried out on the inter-links between biology and psychology. As such, Francis Galton in his seminal investigation of the English scholars found that most of them were first-born, suggesting that older sons get special treatment from the parents that gives them advantage in striving intellectually [@galton1874]. This can be regarded as the first attempt to link cognitive skills and outstanding socio-economic achievement to a wihtin-family dynamics. Psychologist Alfred Adler extended this assumption from intellectual ability to personality, suggesting that first-born, second-born, and later-born children are treated differently by parents, which makes substantial footprint on their non-cognitive traits: while the first-borns enjoy privileges, they also experience the effect of "dethronement" after other children arrive in the family.

These theories triggered scientific discussion on the effect of birth order on personality traits, with the most notable contribution, namely, family-niche theory, arriving in mid-90s and suggested by Sulloway [@sulloway1995; @sulloway1996]. The theory assumes that due to the causal effect of family dynamics, especially when parental resources are limited, the first born is the one benefiting most from the parental investment, having competitive advantage in the number of socio-economic domains. As the order of birth reflects disparity in age and the associated power, this should determine a niche a child occupies in the family dynamics, which is assumed to translate into stable personality differences between siblings. The overall assumptions are that the first-borns will score higher in the non-cognitive traits associated with dominance, whereas later-borns in those linked to adaptability.

Previous research on the matter produced the mixed evidence, both in support and rejection of Sulloway's hypothesis. The analysis also extended the study of gender composition of siblings on personality with some studies highlighting no effect [@dudek2022], while others emphasizing the important role of siblings' gender on personality formation [@golsteyn2020]. Within-family relations also proved to play a role in the development of non-cognitive traits: the analysis that adopted standard skills production function pointed out at the critical role of parental engagement, namely, the engagement of fathers, on the child's socio-economic outcomes throughout the life span [@elkins2020].

Despite the explicit presence of the referenced problem in the psychological body of work, intra-household analysis of siblings is not new to the economic research either. It is well acknowledged by the economists that the "resources are not allocated randomly within households, and that the distribution of resources, pecuniary as well as nonpecuniary, is unequal within the family" [@subramaniam1996]. With that respect, much of the research adopting a within-family design explored how parental and family level decisions could differently affect various socio-economic outcomes in siblings, such as educational expenditure of parents on child education [@kingdon2005], paternal and maternal income on anthropometric outcomes such as height-to-weight [@thomas1990], effect of child's gender on consumption [@subramaniam1996], etc. Akin to psychological studies, birth order effects are also well documented in the economic research when it comes to the differences in the socio-economic outcomes of siblings. The evidence from the within-family studies affirms that birth order affects the nutritional status of a child [@thomas1990a] , investment into human capital [@fors2022], educational disruptions (e.g., school drop outs) under credit constraints [@lafortune2014], and other domains.

With that respect, most of the existing studies predominantly focus on the gaps in the distribution of material assets. However, there is a smaller body of work which explored intra-household disparities between the siblings with respect to non-material aspects, such as disparities in different components of human capital. These studies produced evidence assuring that within-family environment and values, both observed and unobserved, may translate into meaningful gaps in idfferent components of human capital or differences in the patterns of economic behavior. As such, within-family environment could strongly influence acquisition of digital skills by youth going along gender lines, thus amplifying the existing access to STEM jobs and overall inclusive participation in the increasingly digital world for young women [@avanesian2024a]. Another study points out at the effect of birth order and sibling sex composition on cognitive skills in primary school in the developing economy context, where child survival presents constraints. This study concludes that while birth order and sex affect chances of the child survival (favoring boys), if a child reaches primary school age, no effect of sex and birth order persists, wiht girls being actually at consistemnt advantage in cognitive skills if they have elder sisters [@makino2018].

Compelling evidence produced by the experimental study of more than 500 teenagers suggests that family environment factors, such as birth order and sibling sex composition, produce significant effects on economic preferences [@detlefsen2024]. As such, second-born children tend to exhibit lower levels of patience, greater risk-taking behavior, and higher trust. However, these effects are moderated by the sex composition of siblings. The tendency toward increased risk-taking among second-borns is evident primarily in same-sex sibling pairs, whereas effects on trust and trustworthiness are more pronounced in mixed-sex sibling pairs. Notably, patience appears unaffected by the sibling sex composition.

Finally, there is an emerging body of research in economics that explores sibling correlations in non-cognitive skills amongst other components of human capital. For instance, the most fundamental study on the subject matter existing to date postulates that sibling correlations of non-cognitive skills in accordance with Big Five taxonomy range from 0.24 to 0.59, indicating that even for the lowest estimate, one fourth of inequality can be attributed to factors shared by siblings. The study however finds that sibling correlations in cognitive skills are higher than 0.50, indicating that since family characteristics explain more than half of the inequality in cognitive development, they are mor epivotal in this domain [@anger2016].

This is in line with the study based on the Swedish register data that explores sibling correlations for key labir market outcomes, with IQ and non-cognitive skills amongst them [@björklund2012] . While other psychological proxies are used to derive a non-cognitive factor, the study also confirms that sibling correlation for non-cognitive component accounting for 0.32 whereas for IQ 0.47 , family bacground is more important in shaping cognitive ability than personality. The study applies the same prcedure to the sample of mono-zygotic twins that share nearly all of genes finding that correlation in non-cognitive component increases to 0.53. Finally, the analysis of intergenerational similarities produces the smaller results making the authors state that conventional intergenerational approaches severely underestimate the role of family background.

Another study on economic inequality in the United States highlights the role of non-cognitive traits, specifically locus of control and self-esteem, in explaining intra-family variability in socio-economic outcomes [@mazumder2007]. The authors find that these two traits together account for approximately 20% of the observed sibling correlation in earnings, family income, and wages. These findings contribute to a growing body of literature emphasizing that an important part of the intergenerational persistence in economic inequality and status transition is related to non-cognitive factors.

Finally, though there is an abundance of studies on how non-cognitive skills affect education, employment and health, both internationally and based on the russian data. However, the existing studyes based on Russian data dont acknowledge the within-family ocmponent in socio-economic achievement, limiting to control for socio-economic status with some econometric approaches. While there is a significant evidence that non-cognitive skills lead to higher grades at school [@avanesian2022], affect the decisions of continuing education [@rozhkova2021; @rozhkova2021c], and participation in the labor market [@zudina2022; @avanesian2024], previous studies explored these effects in the between-individual manner, without accounting for within-family effects. This approach in a way fails to acknowledge that some families have different likelihood in producing different level of non-cognitive skills in children, and the baseline probability of a certain child to have advanced level of a specific non-cognitive skill is different due to some factors that play out exclusively at the family level. This narrative is somewhat in line with the one proposed by @samuelb1976 that inter-generational transition of socio-economic status occurs primarily through non-cognitive mechanisms, with rich families cultivating in their children independence, openness and creativity, and the poor ones - obedience, agreeableness, and conscientiousness in executing the tasks.

## Data

```{r}
datasummary_skim(sample_data)
```

## Econometric Verification

In order to assess the outlined research questions, the study adopts a within-family design that econometrically is implemented via mixed-effects (multilevel) models. These models allow for 1) accounting for the repeated nature of the data, as the data on non-cognitive skills was collected in 2016 and 2019, and 2) separate within- and between-family variation in estimating the differences in non-cognitive skills between siblings. This is done through the adoption of random intercept terms, which, instead of holding the means constant across the units of the data, allow them to vary across the individuals (for repeated measures) and families. The analysis is implemented in `lme4` package [@lme4] of R, an open-source software for statistical analysis. The statistical significance tests of the effect sizes are produced with the help of `lmerTest` package [@lmerTest].

## Results

### Between- and Within Family Variance in Non-Cognitive Skills

The analysis starts with 5 baseline models, each estimating non-cognitive skills as a function of family and individual id. The model has the following structure:

`m_opns_base <- lmer(NCS  ~ 1 + (1|family_id) + (1|idind), data = ind_fam)`

Where NCS refers to the analyzed non-cognitive skill, representing an intercept-only model. The model partitions the variation in this trait into components attributable to family and individual identifiers. This analysis produces a table of Intra-Class Correlation Coefficients (ICCs), also referred to as Variance Partitioning Coefficients (VPCs). The ICCs represent the proportion of variance in non-cognitive skills explained by specific terms—in this case, family and individual effects. The ICC values provide insight into the degree of clustering within the data. Higher ICCs indicate stronger clustering effects, thereby reinforcing the appropriateness of using a mixed-effects modeling approach to account for these multilevel structures. The results are presented in @tbl-icc-reg.

```{r}
#| label: tbl-icc-reg
#| tbl-cap: Inter-Class Correlation Coefficients of the baseline regression models with random intercepts for individual and family ID. 

gt(icc_baseline) 

```

The findings reveal that all five non-cognitive skills are significantly influenced by family factors, though the degree of influence varies considerably across traits. Agreeableness demonstrates the highest family-level influence, with approximately 41% of its variance attributable to family-level factors. This underscores the critical role of families in shaping interpersonal traits, such as the ability to cooperate, compromise, and collaborate with others.

Openness and conscientiousness exhibit family-level influences of around 32% and 29%, respectively. Openness, linked to curiosity and creativity, reflects how family environments may encourage or limit intellectual exploration, while conscientiousness highlights the familial role in nurturing traits such as diligence, organization, and responsibility. These findings suggest that family dynamics and practices significantly contribute to the development of these traits, potentially through mechanisms such as parental modeling, reinforcement of behaviors, and the establishment of expectations.

Emotional stability follows closely, with 26% of its variance tied to family factors. This finding one quarter of inequality in the traits related to managing emotions and control partially ocucrs due to the shared family characteristics.

In contrast, extraversion shows the lowest family-level influence, with just around 16% of its variance explained by familial factors. This suggests that extraversion is predominantly shaped by individual agency and experiences outside the family, such as social interactions and personal environments.

Notably, the family-level variance incorporates both genetic and shared environmental influences. While the specific mechanisms remain unclear, the relatively high individual-level variances suggest substantial contributions from personal experiences and non-shared environmental factors. Future research could explore the interplay between genetic predispositions and unique environmental contexts in shaping these traits.

These insights are particularly valuable as they indicate that, while genetic factors are not directly observed in this study, their effects are embedded within the family component. This highlights the pivotal role of familial influences in the development of non-cognitive skills and sets the stage for future research to identify the underlying processes at work.



### Socio-Demographic Factors

In this section we extend our analysis to socio-demographic controls that should approximate a family environment while a child was growing. These variables refer to age, age squared, sex, a dummy of being a first-born child, to test the assumption of the privileges associated with being the first recipient of parental investments, as well as a dummy indicating very close relation with the parents (a tfather and mother separately) in the childhood. As the dummy for a close relationship with mothers in childhood did not produce significant effect in any of the five traits and resulted in the poorer model fit, the variable was dropped and the models were re-estimated. This could be potentially explained by the fact that most of respondents reported they were close with their mothers in childhood, which is logical given that mothers have a stronger connection to children in early childhood as the main bearers. With that respecxt, the equation had the following code:

`NCS ~ age + I(age^2) + sex + first_born + relations_with_father + male_siblings_perc + disability + (1 | family_id) + (1 | idind), data = ind_fam`\`

The models are presented in @tbl-socdem-reg.

```{r}
#| label: tbl-socdem-reg
#| tbl-cap: Effects of Socio-Demographic Factors on Non-Cognitive Skills, Mixed Effects Models

modelsummary(soc_dem_models,
             estimate = "{estimate}{stars}",
             gof_omit = "ICC|RMSE", output = "flextable")
```

The output suggests that age does not produce a statistically significant effect on any of the personality traits, which is expected due to 1) the limited age range of the study and 2) the within-family design of the analysis, which separates within- and between-family variation in non-cognitive skills. Interestingly, the effect of being first-born is evident only with respect to openness. This finding aligns with the family-niche theory, which suggests that younger siblings are often more creative and open to new experiences. In this context, being first-born reduces openness to new experiences by 8% of a standard deviation (SD), highlighting the nuanced role of birth order in personality development and in the acquisition of socially desirable skills.

The model also identifies a significant effect of sex, with female youth consistently outperforming male youth across most non-cognitive skills. Specifically, female youth are more open to new experiences, with an effect size of 14% of SD; more conscientious, with an effect size of 23% of SD; and more extraverted and agreeable, with effect sizes of around 14% of SD, for both traits.

The only trait where female siblings show a notable disadvantage compared to their brothers is emotional stability, where females experience a significant negative effect of 24% of SD. This finding suggests that, while female siblings excel in almost all non-cognitive components of human capital, emotional stability may be more susceptible to gendered socialization or environmental influences.

with that respect, gender composition of siblings also plays a role. The inclusion of the share of male siblings (male_siblings_perc) adds an important dimension to understanding inequality in non-cognitive skills, particularly emotional stability. A higher proportion of male siblings in the family is associated with a significant positive effect on emotional stability, increasing it by 21% of SD. This suggests that growing up in a male-dominated sibling environment might foster emotional regulation in youth.


Finally, having a very close relationship with the father during childhood yields significant benefits for agreeableness, with an effect size of 15% of SD, and extraversion, with an effect size of 20% of SD. The strong influence of a close paternal relationship underscores the role of family dynamics in shaping personality traits.

Disability is another important factor in shaping the personality of siblings in adolescence and youth. It has adverse impacts on extraversion and emotional stability, reducing these traits by approximately 19% of SD for both. These findings highlight the complex interplay between familial, demographic, and gender-related factors in the development of non-cognitive skills.

## Decomposition of Variance in Non-Cognitive Skills by Individual Factors

Make a step-by-step addition to the baseline model of the individual level predictors to see what they add to the fixed effects level variance (ie reduction in the ICC for idind).

start with age and age squared model then add sex than add first_born than add relations_with_father than gender composition of siblings than disability

birth order - to include into the baseline model

## Within-Family Effects of Non-Cognitive Skills on Completing Higher Education and Finding a Job

lets see if non-cognitive skills operate separately on the achievement of young people in education and labor market

as two siblings share the same position in socio-economic ladder and starting point in life, in terms of not varying in the parental occupation, education, and household income, the effect of non-cognitive skills on the educational attainment and employment should be similar for both siblings and would absorb the effect of "initial socio-economic conditions" such as observed and unobserved effects of family background that affect the educational attainment and employment of children due to transition of advantage or disadvantage through the inter-generational mechanisms.


## Discussion

One of the very important findings refers to the fact that between-family factors play a critically important role in the individial variation in non-cognitive skills. Depending on the skill, up to 40% of variation in non-cognitive skills can be attributed to family-level factors. With that respect we wanted to test the role of individual effects in the sample of siblings. While sex was a constant predictor of individual differences in the non-cognitive skills of siblings, other variables failed to produce consistent and robust effects.

In line with family-niche theory, first-borns were proved to be less open, as creativity and appreciation of externalities and risk taking is rather developed in later-born children [@paulhus1999]. While first-borns are always perceived by parents as those who have a higher responsibility over the younger siblings and dominate over them, the later-borns develop social strategies that could help them get what they want.

Interestingly, a within-family dynamics with respect to the gender composition of siblings also proved to play role: controlling for a gender of an individual, the study confirms that growing in the families with male-dominant siblings results in higher emotional stability, which potentially indicates at cultural norms and views on gender at the family level that cultivate. Gender composition of siblings in many ways reiterates the effect of gender, however, the individual gender was showing a consistently significant effect throughout, while the effect of gender composition of siblings (with higher values of the ratio indicating dominance of boys and young men), controlling for the individual gender, produced the effects only on emotional stability.

Interesting finding refers to the lack of the effects of obtaining higher education and finding a job. Insignificant effect of employment across all 5 domains of personality, and limited effect of having higher education on openness rather suggest that siblings in higher-education and employment age belonging to the same family do not differ in the baseline probabilities of having higher education or being employed. Further, this age (early 20s) is rather known as the one when personality is rather fixed and all the life transition changes have a little or no effect on personality [@cobb-clark2012; @cobb-clark2013]. Finally, even previous studies highlighted the effect of higher education on non-cognitive skills, pointing out that it rather operates through the exposure to university life, not through the curricular mechanisms [@kassenboehmer2018]. With that respect out findings that being in the process or having obtained higher education results in the increased openness to new experiences is logical.

## Research Limitations

The major limitation of the study refers to inability to separate hereditary mechanisms behind non-cognitive skill formation from those produced by common social environment. Though this could have strong implications on policies for education and labor market on how to nurture these skills from childhood, only a sample of mono-zygotic twins would allow to completely ignore the role of genetic factors in development of non-cognitve skills.

## References
