---
title: "Beyond Wages: The Role of Non-Cognitive Skills in Job Satisfaction of Young Workers"
format:
  tandf-pdf:
    keep-tex: true  
  html: default
author:
  - name: Garen Avanesian
    affiliations:
      - ref: SFedU
    orcid: 0000-0002-0328-637X 
    email: avanesyan@sfedu.ru
  - name: Olesya Shipitko
    affiliations:
      - ref: SFedU
    orcid: 0000-0003-0786-4146
    email: oshipitko@sfedu.ru
  - name: Marina Borovskaya
    affiliations:
      - ref: SFedU
    orcid: 0000-0002-8089-183X 
    email: bma@sfedu.ru
affiliations:
  - id: SFedU
    name: Southern Federal University
    city: Rostov-on-Don
    country: Russia
abstract: | 
  This study estimates the effect of non-cognitive skills on job satisfaction among young adults aged 15–29, using data from the Russian Longitudinal Monitoring Survey (RLMS-HSE) collected in 2016 and 2019. We assess whether non-cognitive skills, approximated by the Big Five Inventory of personality traits (openness, conscientiousness, extraversion, agreeableness, and emotional stability), influence job satisfaction across several key dimensions: overall job satisfaction, satisfaction with working conditions, satisfaction with pay, and satisfaction with career development opportunities. Recognizing the critical role of wages in determining job satisfaction, as well as the potential non-linearity in this relationship, the study aims to isolate the effect of non-cognitive skills on job satisfaction from that of wage-related factors. To address endogeneity arising from variation in wages, a mixed-effects (multilevel) model is employed. The results indicate that extraversion, emotional stability, and conscientiousness have statistically significant positive effects on overall job satisfaction. These effects also vary by wage quintile, with extraversion showing the most pronounced pattern: its effect on job satisfaction is strongest among low-paid workers. Extraversion and conscientiousness also remain significant predictors of satisfaction across the specific domains of working conditions, career development opportunities, and pay. Emotional stability has a positive and significant effect on satisfaction with working conditions and career development opportunities but does not influence pay satisfaction. Finally, domain-specific regressions reveal a significant positive effect of agreeableness on satisfaction with working conditions. The study concludes with practical recommendations for employers on how to enhance job satisfaction among young workers at the early stages of their careers by taking into account their non-cognitive traits.
  
keywords: 
  - job satisfaction 
  - non-cognitive skills 
  - big five personality traits 
  - youth 
  - wages
bibliography: references.bib  
biblio-style: bwl-FU
citeproc: true
csl: "apa-with-abstract.csl"
---

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r wd set up, include=FALSE}
# set up the path

main_folder = "~/Documents/GitHub/NonCognSkillsRLMS"
setwd(main_folder)


```

```{r}
source(file.path(main_folder, "user_profile.R"))
source(file.path(main_folder, "02_codes/libraries.R"))
source(file.path(main_folder, "02_codes/03_JobSatisfNCS/01_data_prep_jobsatisf.R"))
source(file.path(main_folder, "02_codes/03_JobSatisfNCS/02_descriptive_jobsatisf.R"))
source(file.path(main_folder, "02_codes/03_JobSatisfNCS/03_regressions_jobsatisf.R"))
```

# Introduction

The transition from adolescence to adulthood involves a series of critical choices in education, the labor market, and broader life decisions. As youth move from educational systems to the world of work, finding the pathways to decent employment, the integration of this group into the labor market becomes a priority for governments and international development agencies worldwide. Introduced in 2015, the Sustainable Development Goals (SDGs) emphasized the importance of promoting employment and decent jobs for youth within the global development agenda for 2030. However, progress has been limited. In 2015, the global rate of youth aged 15-24 not in education, employment, or training (NEET) stood at 21.3%. By 2024, this figure had only marginally decreased to 20.4%, largely exacerbated by the COVID-19 pandemic, and it is projected to remain stagnant through 2025 and 2026 [@ilo2024]. This highlights the persistent deprivation young people face in gaining the skills needed for inclusive socio-economic participation and accessing opportunities in the labor market.

The Global Employment Trends for Youth 2024 report by the International Labour Organization (ILO), a UN custodian labor agency, underscores the lack of significant progress in creating decent work opportunities for youth. For example, the proportion of young adults aged 25-29 engaged in low-paid jobs varies substantially by region, ranging from 10% in Central and Western Asia to 32% in Sub-Saharan Africa, with Eastern Europe at 15% [@ilo2024youth]. Job security among youth also differs markedly depending on a country’s economic development: while 76% of young people aged 25-29 in high-income countries hold permanent paid positions, this figure drops to just 16% in low-income countries. Alarmingly, the share of youth in temporary work has increased over time, reflecting what the report describes as a global trend toward “the casualization of labour,” which is “a source of increasing anxiety among young people striving to achieve financial independence and transition to adulthood” [@ilo2024youth p. xvii].

The primary challenge in integrating youth into the labor market stems from the mismatch between the supply of young workers entering the labor market annually and the demand for their skills and services. In other words, the number of young people entering labor market substantially exceeds the needs of the employers. This problem persists even in countries experiencing demographic shifts and declining birth rates. Although global youth unemployment rates have declined [@ilo2024youth], these statistics often obscure qualitative issues. In many developing regions, the primary concern is not unemployment but the low quality of available jobs. As noted, “in much of the developing world, the main youth employment issue is the low quality of many of the available jobs more than open unemployment,” with most youth engaged in “self-employment activities or in household-based work, much of which may have quite low returns” and, according to ILO standards, does not qualify as “decent.” [@mckay2018].

The subjective dimension of employment — how satisfied young workers are with their position in the labor market — plays a critical role in shaping different outcomes of youth and young adults, sometimes even beyond the world of work. Low job satisfaction among youth has been linked to counterproductive workplace behavior and substance abuse, including drug use [@mangione1975], alcohol consumption [@kohan2002; @hight2019], and even crime [@chioda2017]. Moreover, job satisfaction significantly impacts retention rates. Satisfied employees are more likely to remain in their roles, whereas dissatisfaction during early career stages often leads to turnover, unemployment, or disengagement from the labor market altogether [@theeffe2011; @chavadi2021; @berber2022; @taris1992; @lehtonen2021], which imposes significant social and economics costs. Poor psychosocial working conditions in a first job adversely affected mental health of labor marker entrants, whereas securing a satisfactory job leads to noticeable improvements in mental health compared to pre-employment status [@milner2016]. All in all, the above underscores that youth represent a pivotal demographic for the labor market dynamics. Job satisfaction is imperative to ensure motivated and effective young workforce that enters labor market.

While common sense suggests that wages should be key determinants of job satisfaction, with better remuneration increasing happiness, the evidence across disciplines is quite mixed. The economic approach to job satisfaction follows this logic: as stated in “the traditional microeconomic models of labour supply, utility from work is related to income one obtains from work and the hours of work that are necessary to secure the given level of income” [@medgyesi2016job]. In other words, economic models frame job satisfaction as the utility derived from labor market choices. Conversely, sociological studies highlight the importance of work values and attitudes in shaping job satisfaction [@kalleberg1977; @kalleberg1993; @kalleberg1983; @wyrwa2020]. In psychology, the relationship between pay and job satisfaction is rather inconclusive [@judge2010]. However, psychological theories provide a more nuanced perspective, offering three distinct approaches to understanding the antecedents of job satisfaction: situational theories, which emphasize the role of objective labor factors (e.g., pay, working hours, and working conditions); dispositional theories, which focus on personality and suggest that job satisfaction is primarily driven by individual traits, with labor conditions playing a relatively minor role; and interactive theories, which combine elements of both approaches [@judge2008job].

Historically, job satisfaction received limited attention from labor economists. The earliest economic investigations, emerging in the late 1970s, primarily focused on the role of relative wages and peer effects on job satisfaction [@hamermesh1977; @freeman1978; @borjas1979; @borjas1979]. These studies were “seen as an economic adaptation of the goal-achievement gap theory” [@lévy-garboua2004] developed by @michalos1980. In recent years, however, economists seeking to explain individual success have increasingly turned their attention to personality. A substantial body of research has since emerged, estimating the effects of personality traits on employment, wages, work productivity, and other socio-economic outcomes beyond labor market (the overview is presented in @borghans2008 and @almlund2011). Economists refer to these traits as non-cognitive skills, emphasizing their malleability during early childhood and the significant role of the social environment, alongside hereditary factors, in shaping them. However, the classification of personality traits as “skills” in economics is primarily based on their productive nature, evidenced by their impact on individual success across multiple domains. To approximate non-cognitive skills, economists commonly rely on the Big Five Inventory, which explains personality variations through five independent factors: openness, conscientiousness, extraversion, agreeableness, and neuroticism (or emotional stability) [@mccrae1987]. Despite their prominence in economic research on human capital and labor market outcomes, the effect of non-cognitive skills on job satisfaction remains largely confined to the psychological domain, which primarily operates on the small samples. However, non-cognitive skills could be particularly critical for young adult workers, who must navigate uncertain job prospects, adjust to mismatches between education and labor market demands, and cope with temporary or precarious employment.

This study addresses this research gap by examining the relationship between non-cognitive skills and job satisfaction among young adults on the Russian labor market. While prior studies have explored this relationship for the general population using nationally representative sample [@zudina2024], no research has specifically focused on young workers transitioning from school to the labor market. Moreover, international research on personality traits and youth job satisfaction remains limited, despite its practical importance for policies aimed at facilitating labor market integration. Young workers face unique challenges, including job insecurity, which increases their likelihood of job turnover and mental health issues [@steenackers2016; @lee2008; @klug2017]. Understanding how non-cognitive skills influence job satisfaction can inform targeted interventions to improve job quality and retention among youth. By addressing subjective employment dimensions, policymakers can design initiatives to reduce disengagement and foster long-term labor market integration.

This study seeks to answer the following research question: How are non-cognitive skills related to overall job satisfaction and specific aspects such as satisfaction with salary, working conditions, and career development opportunities among young adults? To account for the wage-related endogeneity and intergenerational transmission of socio-economic status (SES) that results in youth from the richer families having access to better-paid jobs, this study explores whether the relationship between personality and job satisfaction varies across wage quintiles. Additionally, the study examines whether the U-shaped relationship between wages and job satisfaction identified in prior literature [@clark1996; @brown2009] holds in the Russian context amongst the cohorts of employed young adults. Using data from the Russian Longitudinal Monitoring Survey (RLMS-HSE), which collected non-cognitive skill measures in its 26th (2016) and 28th (2019) waves, this study employs multilevel (mixed-effects) modeling techniques with inverse probability weights to account for sample bias due to employment selection and endogeneity arising from variance in wage [@keil2023; @fuentes2021; @causali2006]. This study represents the first attempt to comprehensively explore the relationship between non-cognitive skills and job satisfaction among young workers in Russia in the context of wage rewards they receive. By leveraging robust methodological approaches and focusing on a demographic at the crossroads of education and employment, it offers valuable insights with implications for labor markets beyond the Russian context.

# Literature Review

## Definition of Job Satisfaction

Scholarly interest to job satisfaction began nearly a century ago. Initially, job satisfaction was primarily studied within the frameworks of applied research in management, sociology, and occupational psychology [@taylor1911; @hoppock1935; @locke1970; @judge2017]. The first general definition of job satisfaction in the literature is attributed to @hoppock1935, who considered job satisfaction as the result of a combination of physiological, psychological, and environmental conditions that enable a person to express a sense of contentment with their work. Later, @locke1970 defined job satisfaction as a value-based response reflecting a subjective assessment of work experience and the extent to which it positively contributes to an individual’s life and development. This value-based response consists of two components: the realization of values (i.e., the extent to which a person is satisfied with the outcome relative to their goals), and the importance of these values within the individual’s personal value hierarchy [@locke1970]. Thus, the significance of specific work aspects mediates the level of satisfaction or dissatisfaction if expectations are not met.

Researchers studying job satisfaction rely on several theoretical approaches that explain the subjective experience of satisfaction. For instance, equity theory focuses on the ratio of effort to outcomes, which employees then compare with those of their peers [@adams1963]. A person may feel dissatisfaction if there is perceived inequity between their own inputs and outcomes relative to those of others. The two-factor theory, developed within the framework of occupational psychology, distinguishes between hygiene factors and motivators as determinants of job satisfaction. A lack of hygiene factors (e.g., salary, working conditions) leads to dissatisfaction, whereas motivational factors (e.g., recognition, achievement) enhance satisfaction but only in the presence of adequate hygiene conditions [@herzberg1971]. Additionally, cognitive dissonance theory of @festinger1957 has been widely applied to assess job satisfaction. For example, the gap between a worker’s current role (e.g., skills or job position) and where they believe they ought to be may generate cognitive dissonance, which in turn affects job satisfaction [@hamermesh2001]. Collectively, these theoretical perspectives have formed the basis for modern research in social and organizational psychology focusing on job satisfaction.

Over the past two decades, the link between job satisfaction and labor productivity has increasingly been studied within the frameworks of organizational psychology and management science. Research findings show that job satisfaction is a critical and at times contentious factor in determining performance—not only of individual employees or specific departments, but also of entire organizations [@jobsati2013].

Numerous empirical studies have explored the relationship between job satisfaction and various personality traits of employees. Findings indicate that emotionally stable individuals are more likely to be satisfied with their jobs [@hoppock1935; @fisher1931], as are extraverts who engage more actively with colleagues and clients, deriving greater enjoyment from their work [@watson1997]. Moreover, job satisfaction has been linked to general self-esteem, which, together with the perceived complexity of the work, significantly determines satisfaction levels [@judge2000].

As noted earlier, such psychological studies have typically been based on small, occupation-specific samples involving professions that require specialized competencies. For example, research has shown that information technology (IT) professionals are generally satisfied with interpersonal relationships at work and with the content and meaning of their jobs, yet report dissatisfaction with pay and career advancement [@okpara2004]. Another study conducted on a sample of nurses, which accounted for various demographic characteristics (age, education, marital status, work experience, and qualifications), found that these factors mediate dissatisfaction related to managerial relations and professional development opportunities [@yaktin2003]. While healthcare personnel report satisfaction with their work outcomes, they are similarly dissatisfied with compensation, echoing the findings in the IT sector. These studies highlight that domain-specific aspects of job satisfaction can be aggregated into a broader construct of overall career and organizational satisfaction. Thus, research shows that industry-specific characteristics significantly shape the subjective perception of job satisfaction.

An exception in terms of sample size is a study conducted on a large sample of general education teachers (over 3,000 respondents) [@gil-flores2017]. Using hierarchical linear models, researchers assessed the importance of teachers’ personal characteristics and school environments as predictors of job satisfaction. The key personal factors associated with job satisfaction included perceived work effectiveness, classroom management skills, and demographic variables such as age, gender, and experience. More experienced female teachers reported higher job satisfaction compared to younger male teachers. In addition, full-time staff and those with long tenure at a single school were more satisfied than part-time employees. At the organizational level, interpersonal relationships with students and colleagues emerged as significant predictors of job satisfaction.

## Empirical Insights on Big Five Personality Traits and Job Satisfaction

From an empirical standpoint, the Big Five personality model has proven to be a reliable predictor in behavioral research. Becoming a mainstream in the study of personality effects on a wide arraw of socio-economic outcomes, this model has been widely used not only in psychology, but also in other disciplines such as economics or sociology. Studies incorporating this model in the context of job satisfaction offer a range of insights, although findings vary. A consistent conclusion across the literature is the empirical validity of the Big Five as a framework for understanding job satisfaction [@cooper2014]. However, despite this theoretical robustness, there is a relative paucity of studies that explore the practical implications of personality traits in organizational settings. More specifically, there remains a gap in research that jointly considers both organizational context and age-specific cohorts. This omission is particularly critical given that, in the long term, young individuals are likely to face challenges related to satisfaction in their professional lives.

Most studies find that general job satisfaction is significantly and positively associated with extraversion. Individuals who are open, outgoing, and sociable tend to be more satisfied with various aspects of their work, particularly interpersonal relations and labor renumeration [@furnham1986]. Interpersonal skills approximated by extraversion serve as a source of support, ease adaptation in complex work environments, and contribute positively to the overall organizational climate. Interestingly, the link between extraversion and job satisfaction is stronger when individuals assess their own performance positively. Conversely, when assessments come from external sources, the relationship weakens or becomes non-significant [@judge2008]. This finding has notable implications for the organizational management, suggesting that companies may benefit from aligning incentive structures with intrinsic motivational factors. The contribution of extraversion is particularly important when the job requires frequent interaction with others. However, the nature and strength of this relationship may vary by occupational role [@furnham1986]. Notably, studies from Asia reinforce this conclusion and further suggest that cultural differences do not moderate the relationship between extraversion (or other Big Five traits) and job satisfaction [@templer2011].

Conscientiousness is commonly associated with a strong desire to perform tasks efficiently and effectively. As such, it is often positively reinforced within organizational settings. However, its link to job satisfaction is typically moderate in magnitude [@judge2008; @templer2011]. While conscientiousness is a stable personality trait that can indirectly support career advancement, its effect on job satisfaction is not always pronounced. Some researchers suggest that conscientious individuals may benefit from favorable working conditions, such as better office environments, which can, in turn, enhance satisfaction. Among younger workers, positive work conditions may help compensate for lower material rewards due to limited experience or skill gaps. Interestingly, not all studies have found a direct link between conscientiousness and job satisfaction. For instance, @seibert2001 found no significant relationship, positing that conscientiousness may not necessarily translate into career advancement or job satisfaction. In some occupational settings, job tasks may be routine and stable, reducing the impact of conscientiousness. Therefore, conscientiousness may only predict job satisfaction meaningfully when considered in conjunction with other personality traits.

Most studies agree that neuroticism is inversely associated with job satisfaction and emphasize the dispositional nature of this relationship. In other words, an individual’s tendency to experience dissatisfaction is influenced by their emotional stability, or lack thereof. Among the Big Five traits, neuroticism has the strongest and most consistently negative association with job satisfaction [@judge2002]. Although this is generally expected, its implications vary across research contexts. Individuals high in neuroticism are more prone to negative emotions, making them likely to feel dissatisfied regardless of whether the source is pay, coworkers, or job tasks [@seibert2001]. Such individuals tend to perceive work situations pessimistically, and no amount of organizational intervention may significantly alter their job satisfaction. Both primary studies and meta-analyses support the conclusion that neuroticism is a strong negative predictor of both job satisfaction and career advancement [@judge2010].

Findings on the role of agreeableness in job satisfaction are mixed. Some studies exclude it as a significant predictor altogether [@sutin2009], while others find that agreeableness may negatively mediate job satisfaction in roles involving high levels of social interaction [@seibert2001]. Agreeable individuals may be less likely to assert themselves, more inclined to defer to others, and more focused on fulfilling others’ goals, all of which can diminish their own job satisfaction. Conversely, more recent studies suggest that agreeableness may positively influence career advancement by facilitating strong relationships with colleagues and supervisors [@gil-flores2017]. Agreeable individuals often contribute to a positive workplace climate, which may enhance their own sense of satisfaction. However, meta-analyses generally show that agreeableness is not a reliable predictor of job satisfaction—particularly when compared to traits like neuroticism, extraversion, and conscientiousness [@judge2002].

Finally, research linking openness to job satisfaction typically identifies a weak or negative correlation, particularly when satisfaction is assessed externally [@judge2008]. Individuals high in openness tend to thrive in roles that involve creative projects and innovation. Organizational strategies that support job satisfaction for these employees may involve delegating greater autonomy and allowing for project ownership. Among younger workers, this approach can be especially meaningful, providing not only job satisfaction but also opportunities for self-actualization. One notable finding by @seibert2001 is that individuals with high levels of openness often earn lower wages; however, this does not appear to reduce their job satisfaction. While the authors do not offer an explanation, it is plausible that demographic factors, such as age, gender, or socioeconomic status, play a role. At early career stages, young adults may prioritize experience and self-development over financial rewards. Given that both earnings and experience are partial components of job satisfaction, future research should explore this relationship in more detail.

Contrary to popular assumptions, even substantial differences in earnings do not necessarily determine job satisfaction. Rather, overall satisfaction is more closely tied to non-cognitive traits. Recognizing this, organizations can design targeted motivational and support systems that leverage employees’ personalities to enhance both individual well-being and organizational effectiveness.

# Data

The present study utilizes data from the RLMS-HSE survey waves conducted in 2016 and 2019, which collected information on non-cognitive skills. The RLMS-HSE is a nationally representative longitudinal survey of the Russian population, conducted since 1994. It gathers extensive data on various dimensions of individuals’ lives, including socio-economic status, health, education, and employment. The survey sample is drawn from 38 regions across the country, and regional differences are controlled for in the empirical analysis. From the broader pool of individuals who completed the adult questionnaire and responded to the Big Five personality traits module, the current study narrows the focus to youth aged 15–29 years, in accordance with the definition of the school-to-work transition. Given that job satisfaction is reported starting from the age of 16, the lower age bound was adjusted accordingly, as respondents must be employed to answer questions on job satisfaction.

The final sample pools observations from both survey waves, resulting in a dataset comprising `r nrow(youth_job_satisf)` observations, of which `r sum(youth_job_satisf$year == "2016")` originate from the 2016 wave and `r sum(youth_job_satisf$year == "2019")` from the 2019 wave. In total, `r length(unique(youth_job_satisf$idind))` unique individuals are represented in the dataset. Descriptive statistics of the analytical sample are presented in @tbl-descr-sample.

```{r}
#| label: tbl-descr-sample
#| tbl-cap: Descriptive Statistics of the Sample

summary_stats_js

```

## Output Variables

Overall job satisfaction refers to the main variable of the interest in the current study. It is based on the question “Overall, how satisfied are you with your job?” The response options range from 1 (very satisfied) to 5 (very unsatisfied). The survey also includes questions about job satisfaction with working conditions, pay, and career development opportunities. The response options for these questions are similar to the overall job satisfaction question. As we assume high correlation between overall job satisfaction and it's specific domains, the focus of the analysis is on the overall job satisfaction.

However, we also explore the specific domains of job satisfaction to provide a more comprehensive understanding of the relationship between non-cognitive skills and job satisfaction. In other words, while overall job satisfaction is the primary outcome of interest, analyzing its specific domains—satisfaction with working conditions, pay, and career development opportunities—offers important analytical leverage. The survey items are designed to capture different facets of job satisfaction, allowing for a nuanced analysis of how non-cognitive skills influence various aspects of the intrinsic career success. Further, estimating separate models for each domain allows us to test the robustness and consistency of the effects of non-cognitive skills across different facets of the individual job experiences. This domain-specific approach helps uncover whether particular traits are more strongly associated with certain aspects of job satisfaction, thereby offering a more nuanced understanding of how non-cognitive skills shape subjective labor market outcomes.

The Spearman rank correlation coefficients between overall job satisfaction and its domain-specific components are presented in @fig-corr-jobsatisf. The results indicate that all job satisfaction domains exhibit moderately strong to strong positive associations. Specifically, the highest correlation is observed between overall job satisfaction and satisfaction with working conditions ($\rho$ = 0.76), suggesting that for young individuals, perceptions of labor conditions are a key determinant of general job satisfaction. It is followed by the moderately strong correlation between overall satisfaction and satisfaction with career development opportunities ($\rho$ = 0.61), implying that individuals who perceive favorable prospects for professional growth are are also more likely to be overall satisfied with the job. Other correlations, ranging from 0.54 to 0.57, indicate moderately strong associations among the remaining domains, suggesting that while each facet contributes to the overall experience, they also reflect relatively distinct yet interconnected aspects of job satisfaction.

The descriptive statistics of the overall job satisfaction and its specific domains are presented in @tbl-descr-sample. In order to make the analysis more intuitive to interpret, we converted the job satisfaction variables into binary measures where 1 denotes categories that refer to being either very satisfied or satisfied, and 0 otherwise.

```{r}
#| label: fig-corr-jobsatisf
#| fig-cap: Correlation Matrix of the Overall Job Satisfaction and its Specific Domains

js_corr_matrix

```

## Input Variables

The input variables that were inserted to the model include, in addition to socio-demographic controls such as sex, age, region, and area of residence, also refer to highest level of education completed, occupation, wage quintile group, hours of work per week, and non-cognitive skills in accordance with the Big Five personality traits model.

# Methodology

Before analyzing the effect of non-cognitive skills on job satisfaction, the study first examines the relationship between hourly wages and job satisfaction among youth. To account for potential non-linearities in this relationship, we employ a generalized additive model (GAM) [@hastie2017a]. GAMs refer to the regression models that offer flexibility by fitting non-parametric cubic spline terms, which effectively capture curvilinear relationships between hourly wages and job satisfaction. While log-transformation of wages is a common approach to address non-linearity, it assumes a uniform diminishing marginal effect. This approach, however, may fail to identify potential variation in the effects of wages on job satisfaction after the point of satiation typical for logarithmic functions. While preserving the logarithmic transformation of wages to address skewness and minimize the influence of outliers, we apply a non-parametric smoothing term to the log of hourly wages to capture potential non-linearities and explore how the relationship evolves across the entire wage distribution. This approach allows us to identify patterns, such as points of diminishing returns, that might otherwise remain obscured by simpler transformations. The analysis was conducted using the `mgcv` package in `R` [@mgcv]. A detailed discussion of the GAM methodology is beyond the scope of this paper but is available in @wood2006.

Following this preliminary analysis, we focus on the primary research question: the effect of non-cognitive skills on job satisfaction. We hypothesize that this effect may vary across different wage levels, emphasizing the need to understand the relationship between the wages and job satisfaction in the previous step. To test the effects of non-cognitive skills on job satisfaction, we adopt a multilevel (mixed-effects) modeling approach. This approach is well-suited for our data as it accounts for multiple sources of endogenous variation in the outcome variable through random intercepts and slopes.

As our dependent variables refer to binary measures, we draw a linear probability multilevel model, which holds a number of advantages in comparison to the logit function. First, it allows for a more straightforward interpretation of the coefficients, as they represent the change in probability associated with a one-unit change in the predictor variable (standard deviations in case of non-cognitive skills). Second, it is computationally less intensive than the logit model, especially when dealing with large datasets or complex models that adopt multiple random terms, like in our case.

Given the unbalanced panel dataset collected in 2016 and 2019, the model accounts for the repeated nature of the observations by including random intercepts for individual IDs. This specification addresses intra-individual dependencies in the data. Regional disparities in labor market access and outcomes are well-documented; thus, we incorporate random intercepts for regions to account for spatial heterogeneity in job satisfaction.

Further, we recognize the complex interplay between job satisfaction, non-cognitive skills, and wages. We posit that wages are a significant driver of variation in job satisfaction on the one hand, and acknowledge that personality traits are strong predictors of individual earnings on the other hand [@edin2022; @lindqvist2011; @collischon2019]. To isolate the effect of non-cognitive skills on job satisfaction, the model estimates their impact within each wage quintile by introducing random slope terms for non-cognitive skills. First, this specification allows us to explore the effect of non-cognitive skills on job satisfaction without the bias that wages could introduce into this relationship due to endogeneity. Second, the proposed approach tests if the effect of non-cognitive skills on job satisfaction is heterogeneous with respect to different wage levels. The mixed-effects models are calculated with the help of `lme4` package in `R` [@bates2015]. As the package does not support the calculation of statistical significance for the predictors, we employ the `lmerTest` package to obtain p-values [@kuznetsova2017].

Finally, another methodological detail refers to the sample bias that occurs due to the self-selection into employment. In order to overcome this limitation, we adopt inverse probability weights based on propensity scores that correct for the probability of employment with the `WeightIt` package in `R` [@WeightIt]. This approach allows us to account for the potential bias arising from the non-random selection of individuals into employment and ensures that the results are generalizable to the broader population of young adults in Russia. The weights are calculated using a logistic regression model that predicts the probability of employment as a function of exogenous factors such as age, sex, level of education, region, and area of residence. The weights are then incorporated into the mixed-effects models to adjust for the sample bias and ensure the robustness of the results.

# Results

## Relationship Between Wages and Job Satisfaction

The results of the GAM regression presented at @fig-wage-jobsatisf affirm that the relationship between the log of the hourly wage and job satisfaction is curvilinear. For the lower levels of wages, the degrees of job satisfaction are also below the average effect of the model (i.e., 0 on the y axis). However, with the increase in wage, they also grow, than reach the satiation point between the 7 and 8 of the log of the hourly wage, and then dramatically fall down. The concave relationship means that after a certain point, higher wage does not lead to the higher job satisfaction, implying that some other factors come into the play. However, this also means that ignoring this curvilinearity is critical in terms of understanding the interplay between wages, job satisfaction, and non-cognitive skills. It guides the analysis to the next step where the key sources of variation in job satisfaction are assessed, including the investigation of how much variation in the overall job satisfaction can be attributed to the wage differentials.

```{r}
#| label: fig-wage-jobsatisf
#| fig-cap: The Effect of Log of the Hourly Wage on the Overall Job Satisfaction, Results of the GAM Regression
#| fig-width: 6
#| fig-height: 5

plot(gam_job_satisf, col = "blue")

```

## Sources of Variation in Job Satisfaction

We start the multilevel analysis by exploring the sources of variation in non-cognitive skills. The baseline mixed effect model, without the predictors in the fixed part of the model, includes random intercept terms for individual id, region, and quintile groups of the hourly wage. The results of the Intraclass Correlation Coefficients are presented in @tbl-icc-baseline. The output suggests that regional factors explain almost 2% of variance in job satisfaction. Differences due to occupational factors account for 3.4% of variance in job satisfaction, whereas differences across quintiles of the hourly wage explain 3.3%. Finally, almost 23% of variation in job satisfaction can be attributed to all other factors between individuals.

While through the random intercept terms we can estimate the shares of variance explained by each of the terms in the outcome variable, it also makes sense to estimate what share of variance in job satisfaction can be attributed to non-cognitive skills. This can be done by exploring the difference in the Marginal R-squared value between the model 1 that predicts job satisfaction based on socio-economic characteristics and the model supplemented by the non-cognitive skill measures. In the context of mixed-effects models, marginal R-squared indicates the share of variance explained by the fixed part of the model, i.e., predictors that do not vary. This is opposed to the Intraclass Correlation Coefficients that returns variance components of random, i.e., varying effects. These coefficients are presented in @tbl-mem-ncs. As such, the difference between marginal R-squared of the first model that includes only socio-economic characteristics, and the one supplemented by non-cognitive skill measures, accounts for 1.1%. In other words, this is the contribution of non-cognitive skills into the explained variance of job satisfaction captured through the fixed effects.

```{r}
#| label: tbl-icc-baseline
#| tbl-cap: 'Intraclass Correlation Coefficients of the Baseline Mixed-Effects Model, Job Satisfaction by the Random Intercept Terms of Individual ID, Region, and Hourly Wage'
#| 

icc(m0_satisf, by_group = TRUE) %>%
  as.data.frame() %>%
  mutate(Group = case_when(Group == "idind" ~ "Individual ID",
                           Group == "region" ~ "Region",
                           Group == "occupation" ~ "Occupation",
                           Group == "hourly_wage_quintile" ~ "Wage Quintile")) %>%
  rename(`Random Term` = Group) %>%
  kable(format = "latex", booktabs = TRUE, digits = 3)
```

## Non-Cognitive Skills and Job Satisfaction

The analysis of the effects of socio-demographic predictors included in the carried out mixed-effects regressions summarized in @tbl-mem-ncs suggests that non of them has a statistically significant effect on job satisfaction when it comes to the youth cohort. The age and age squared coefficients suggests that within the studied age brackets, age does not affect job satisfaction. Further, there is no statistically significant effect of sex on job satisfaction. Same applies to education and area of residence. However, both model with socio-economic variables and the one supplemented by non-cognitive skills, identified the statistically significant effects of wages and hours of work. In line with the economic approach wages have positive effect and hours of work have negative effect on job satisfaction, as individuals aim to maximize their utilities with the minimum input. As such, log term of hourly wages positively affects job satisfaction, with the increase in the hourly wage on 1% resulting in the 9.5% of probability of being satisfied with the job. Interestingly, the effect of log of the hourly wages slightly redcues in the model supplemented with non-cognitive skills, suggesting that non-cognitive skills may partially mediate the effect of wages on job satisfaction, and the effect initially attributed to the wages could arise due to non-cognitive skills. On the opposite, increase in the hours of work per week on 10 hours decreases the probability of job satisfaction by 1% (the effect is marginally significant at p\<0.1), and this effect does not change in socio-economic and non-cognitive skills model.

When it comes to non-cognitive skills, out of 5 characteristics, the model identified positive and statistically significant effect of three, namely, conscientiousness, extraversion, and emotional stability. Extraversion produces the highest effect in magnitude, with the increase on 1 standard deviation in this trait resulting in the 4% increase in the probability of being satisfied with the job. It is followed by emotional stability, which increases job satisfaction likelihood by almost 3%, and conscientiousness which results in a 2.6% higher probability of being overall satisfied with the job.

```{r}
#| label: tbl-mem-ncs
#| tbl-cap: 'Mixed-Effects Model Regressions of Job Satisfaction: 1) Socio-Economic Factors, 2) Socio-Economic Factors and Non-Cognitive Skills, 3) Model with Random Slopes of Non-Cognitive Skills by Wage Wuintile Groups' 

models_job_satisf

```

## Does the Effect of Non-Cognitive Skills Vary by Wage Quantile?

Exploring the interplay between the job satisfaction, wages, and non-cognitive skills, we built the third model that excludes the effect of log of the hourly wage but instead takes the random slope of non-cognitive skills by hourly wage quintile groups. While the coefficients of fixed terms can be found in the third column of @tbl-mem-ncs, the results of the random part fo the model are plotted on @fig-wages-ncs. As the figure suggests, the most extraversion establishes the most interesting pattern, suggesting that the lower is the wage, the higher is the effect of extaversion on job satisfaction. As such, accounting for more than a 5% increase in the probability of being satisfied with the job for the youth from the bottom 20% by hourly wage, its coefficient gradually decreases along the wage ladder, dropping to 2.38% increase for the youth from the 5th quintile of hourly wage. Somewhat similar pattern is observed for conscientiousness, as its effect on job satisfaction is rather higher for the lower wage quintile groups, though variation is not as substantial. It rather overall can be said that the more conscientious a person is, irrespective of their wage level, the more satisfied with the job they are. However, the highest effect of conscientiousness on job satisfaction is observed amongst the youth from the second wage quintile (3.39% increase in probability), whereas the lowest for the youth from the top quintile (2.02% increase). Interestingly, the effect of emotional stability is the most pronounced for the youth from the second quintile of hourly wage, i.e., those not at the bottom, yet below the median wage, and accounts for almost 5% increase in the probability of being overall satisfied with the job. It is followed by the coefficient of 2.8% of increase in probability of being satisfied with the job amongst youth from the 4th quintile, i.e., those not with the top pay, yet above the median level.

```{r}
#| label: fig-wages-ncs
#| fig-cap: Random Slope Coefficients of Non-Cognitive Skills on Job Satisfaction, Results of Mixed-Effects Regression
#| fig-width: 5
#| fig-height: 5

print(plot_wages_ncs)
```

## The Effects of Non-Cognitive Skills on the Domain-Specific Aspects of Job Satisfaction

On the final stage of the analysis, we explored the effect of non-cognitive skills on three other domains of job satisfaction: satisfaction with career development opportunities, satisfaction with working conditions, and satisfaction with wages. The mixed-effects models are presented in @tbl-mem-ncs2.

```{r}
#| label: tbl-mem-ncs2
#| tbl-cap: 'Mixed-Effects Model Regressions of Satisfaction with: 1) Career Development Opportunities, 2) Working Conditions, 3) Wages' 

models_satisf2
```

The results obtained across the three domains of job satisfaction largely corroborate the findings of the overall job satisfaction model. Conscientiousness and extraversion demonstrate positive and statistically significant effects across all three domains examined. The effect of conscientiousness is most pronounced in relation to satisfaction with career development opportunities, contributing to a 4.5% increase in the probability of being satisfied with opportunities for career advancement. Additionally, conscientiousness is associated with a 3.2% increase in the probability of being satisfied with wages and a 2.2% increase in satisfaction with working conditions.

Similarly, the effect of extraversion is most substantial for satisfaction with career development opportunities, resulting in a 4.3% increase in the probability of being satisfied with opportunities for professional growth. Extraversion is also linked to a 3.0% increase in satisfaction with working conditions and a 2.8% increase in satisfaction with wages.

Emotional stability exerts a statistically significant positive effect on satisfaction with both career development opportunities and working conditions, increasing the probability of satisfaction by 2.6% and 2.4%, respectively. Finally, the analysis reveals that individuals with higher levels of agreeableness are more likely to report satisfaction with their working conditions, with a 2.0% increase in the probability of being satisfied.

# Research Limitations

The study is not without limitations. First, the analysis relies on self-reported measures of job satisfaction, which may be subject to biases and inaccuracies. The original variables represents an ordinal scale consisting only of 5 items each. Further, only one question is used to measure each concept. Together, these two aspects prevent from aggregating the measure and carrying out the analysis approaching it in a quasi-continuous manner that better captures variation in job satisfaction. Instead, the analysis relies on the binary measure of job satisfaction, which may not fully capture the nuances of individuals' experiences.

Furthermore, although the study controls for variation in the baseline probability of job satisfaction due to occupation through the random intercept terms, it does not account for the influence of contextual and organizational factors, such as work-life balance, job security, and organizational culture, all of which may significantly shape employees’ job satisfaction. These variables could act as confounders or moderators, potentially amplifying or mitigating the impact of non-cognitive skills. Their omission limits the ability to fully understand the interplay of influences on job satisfaction, highlighting a valuable avenue for future research to explore more nuanced, interactional models.

# Discussion

The findings of this study contribute to the growing body of literature on the role of non-cognitive skills in shaping individuals’ socio-economic outcomes in the Russian context. While previous evidence on the relationship between non-cognitive skills and extrinsic antecedents of labor market success in Russia is robust and well-established [@rozhkova2019; @maksimova2019; @zudina2022; @avanesian2024], the present study extends this understanding by demonstrating that non-cognitive skills also play a significant role in intrinsic antecedents of labor market outcomes such as job satisfaction. Notably, even within international research, this topic has received limited attention from economists, having been primarily examined from a psychological perspective focusing on the relationship between personality traits and job satisfaction.

The results of this study largely align with prior psychological research, albeit based on smaller samples without national representativity. Meta-analytical studies underscore the strong effects of extraversion, neuroticism/emotional stability, and conscientiousness. For instance, the seminal work of @judge2002, which synthesized 334 correlations across 16 independent samples, identifies these three traits as the most influential, with the effects of extraversion and emotional stability being generalized across studies.

Moreover, our findings are consistent with those of @zudina2024, who, using nationally representative data from the general working population, concludes that extraversion, emotional stability, and conscientiousness are universally associated with higher subjective job rewards in the Russian labor market. Interestingly, the fact that non-cognitive skills affect job satisfaction among youth similarly to their effects among the general working population differs from what is observed with extrinsic job outcomes, such as wages. Existing research suggests that the impact of non-cognitive skills on wages evolves across the life course as individuals progress through various stages of labor market participation [@avanesian2025]. This divergence implies that while the economic returns to non-cognitive skills on the Russian labor market may depend on career stage and accumulated experience, their contribution to subjective job rewards appears to be more stable and immediate. In other words, non-cognitive skills such as extraversion, emotional stability, and conscientiousness may enhance young workers’ capacity to find satisfaction in their current roles, thus playing a critical role in fostering early workplace engagement and retention.

# Recommendations

Research findings point to essential measures that organizations should adopt to enhance their effectiveness when engaging young employees. In the search for these determinants, our study turned to the concept of non-cognitive abilities as a broader aspect of personality, thus taking into account factors that mediate both performance and satisfaction at work.

It is important to consider the identified relationship between wages and job satisfaction. The GAM results demonstrate a curvilinear relationship between wages and job satisfaction, with a plateau and decline in satisfaction at higher income levels. Increasing economic incentives is not always feasible and, moreover, as per data, is not sufficient. While such measures may yield short-term benefits, their effect diminishes once the young workers progress in their career to more advanced and better-paid positions. In addition to equitable labor remuneration, organizations should pay attention to alternative determinants of job satisfaction, such as career guidance, feedback, recognition, and autonomy - aspects that are crucial to enhance productivity and foster a positive psychosocial climate in the workplace.

As our findings confirmed that extraversion, emotional stability, and conscientiousness have a significant and positive impact on job satisfaction, employers must recognize the importance of non-cognitive skills and invest in their development, especially among young employees. It is therefore advisable to design and implement programs aimed at developing these non-cognitive skills, particularly at the early stages of a professional career. Substantial benefits can be achieved through the use of modern training technologies focused on enhancing interpersonal skills, emotional intelligence, and self-management skills.

Extraversion emerged as the strongest predictor of job satisfaction among young workers, especially those in the lowest wage quintiles. This implies that interpersonal and emotionally expressive traits become especially critical in environments where financial incentives are limited. It also necessitates the use of more differentiated strategies for motivating and developing young personnel. Satisfied employees are generally more sociable, energetic, cooperative, and tend to amplify their work outcomes. High levels of extraversion indicate a capacity for quickly establishing social contacts and a desire to receive support from the organization. This can be seen as a form of long-term retention of the young workforce by the organization.

Job satisfaction is influenced not only by non-cognitive skills but also by working conditions and attention to employees’ career trajectories. In other words, recognizing, first, the complexity of this phenomenon, and, second, the fact that personality affects the way people are satisfied with the different aspects of their job, should be taken into account by the human resources professionals. Since agreeable individuals tend to be more satisfied with the working conditions, improving physical and social working environments for them is essential to make sure they are not satisfied by having less. Same applies to enhancing role clarity and promotion pathways for emotionally stable employees. Together, this may yield better outcomes than generalized engagement strategies that, based on the results of our study, would produce a response only in the workers with specific set of traits.

Consequently, organizations should strive to create a supportive and inclusive work environment. This can be achieved through transparent human resource policies. Young professionals should be well informed about opportunities for career advancement and personal growth. Access to upskilling and professional training should be unambiguous and reinforced by the organization’s commitment to balanced workloads and adequate rest, including company-supported initiatives that facilitate work-life balance.

In light of our findings, particular attention should be paid to the development of conscientious employees, who contribute to organizational goals through planning and advanced executive function. A comprehensive approach to understanding job satisfaction, one that encompasses wage levels, the quality of communication between staff and leadership, interest in the work itself, corporate culture, and other relevant aspects, allows for the identification of more effective personnel strategies. Implementing a monitoring system, such as employee satisfaction surveys, to track job satisfaction among young work force would be beneficial, helping organizations to better integrate them and increase their productivity.

The results underscore the need for targeted strategies to engage and retain young professionals. Programs promoting fast-track career advancement through participation in complex projects, along with a positive moral and workplace environment, can foster a strong bond between staff and organization. Structuring work around project-based models helps to maintain engagement. Recognizing and rewarding achievements of the young workforce can sustain their motivation and reinforce their sense of self-worth.

Given that non-cognitive skills influence satisfaction even before material rewards accumulate, the early career period represents a critical window for shaping positive workplace experiences. Cultivating inclusive, transparent, and socially supportive environments for young professionals, especially those in lower-paid or unstable roles, can enable organizations to build a healthier work environment, reduce staff turnover, and improve overall labor productivity.

# Conclusion

This study examined the impact of non-cognitive skills on job satisfaction among Russian youth aged 15–29, using nationally representative longitudinal data from RLMS-HSE collected in 2016 and 2019. By leveraging mixed-effects models that accounts for wage-related variation through the random slope term, the analysis isolated the unique contribution of Big Five personality traits to both overall and domain-specific job satisfaction.

Our findings reveal that extraversion, conscientiousness, and emotional stability significantly enhance job satisfaction, even after controlling for wages and other socio-demographic factors. Importantly, the effect of these traits varies across wage quintiles, with extraversion exerting the strongest influence among low-paid workers. Furthermore, the relevance of each trait depends on the specific domain of job satisfaction, underscoring the multidimensional nature of job satisfaction as a phenomenon.

These results extend the existing literature in labor economics by highlighting that non-cognitive skills not only influence labor market outcomes such as employment status or wages (which are extrinsic in nature), but also shape intrinsic job success by producing a subjective reward. While economic incentives remain important, the study demonstrates that non-cognitive skills offer an additional pathway for improving job satisfaction and job retention among young workers — a group particularly vulnerable to early career disengagement.

From a policy and organizational perspective, the findings underscore the importance of integrating non-cognitive skill development into youth employment strategies. Interventions aimed at fostering emotional resilience, interpersonal skills, and conscientious work habits may yield substantial returns, especially when tailored to wage levels and specific workplace environments.

By bringing together personality psychology and labor economics, this study contributes new empirical insights to the body of work on youth employment and offers practical guidance for enhancing job satisfaction through non-monetary means. In doing so, it also affirms the critical role of personality in shaping meaningful and sustainable labor market participation.

# References
